### 2023-10-10 chapters/discussion.tex
Updates to discussion

During the three years since the start of this project, the rapid development of machine learning that started about a decade ago with the ImageNet competation and the work of \textcite{krizhevsky_imagenet_2012} all but slowed down. 
In this section we will briefly revisit selected chapters of the thesis and provide some additional thoughts and discussion.

\section{Speech representation learning}
%
In this thesis we have studied two different approaches to learning speech representations. Variational autoencoders in \cref{chp:paper-hierarchical,chp:paper-modelagnostic,chp:paper-benchmarking} and self"=supervised methods in \cref{chp:paper-review}. 
We found that the probabilistic formulation of variational autoencoders provide benefits for their application to uncertainty quantification but that they are generally inferior to self"=supervised methods when it comes to performance on downstream tasks such as phoneme recognition (see e.g. \cref{tab: phoneme recognition (PER), tab:unsupervisedASR}).

VAEs are good at uncertainty quantification (let's at least say that they are).

SSL methods are very good at representation learning as reflected by their dominance on downstream task performance in for text and speech.

It seems like we cannot have a model that has both a principled probabilistic representation of uncertainty and the ability to learn rich representations that are useful for downstream applications.

@@

One way to improve the representations learned in VAEs is via semi-supervised learning. 
Here a small number of curated labels are used to inform and enrich the patterns that can be learned from a large unlabelled data set. 
VAEs are strong for semi-supervised learning but there are other strong alternatives too \cite{jiang_speech_2021}. Additionally, semi-supervised learning is more constraining than is self-supervised methods that rely only on pretraining; a single foundation model can be fine-tuned at relatively low cost for a large number of downstream tasks, while semi-supervised models must learn all relevant downstream tasks while simultaneously learning from the unlabelled data, which is expensive.

@@

(layers )

Masked pretrained as maximum likelihood

@@

\section{\Cref{chp:paper-retrospective} revisited: Calibration}
The work in \textcite{wenstrup_retrospective_2023} focuses on the predictive performance of the ensemble model and an analysis of feature importance, but does not explicitly consider uncertainty estimation. 
As we discussed in \cref{subsec:model-calibration}, such a model can be calibrated to predict probabilities that are aligned with the empirical probability of the model being correct on some validation set.

We compute the calibration curve for the stroke recognition ensemble by sorting the predicted probabilities on test set into a number of bins spanning the range from zero to one. For each bin, we compute the fraction of examples for which the model predicted correctly and then plot these pairs of mean predicted probabilities and correctness fractions. A perfectly calibrated model would have the same fraction of correct predictions in any given bin, as that bin's mean value. 
In \cref{fig_discussion:retrospective-paper-calibration-curve-of-uncalibrated-model}, we plot the calibration curve for the uncalibrated stroke recognition ensemble and the individual models along with a histogram of its predicted probabilities. The miscalibration issue that we previously discussed is clearly visible as a strong overconfidence.

\begin{figure}
    \begin{subfigure}[c]{0.48\columnwidth}
        \centering
        \includegraphics[width=1\columnwidth]{paper_retrospective_calibration_plots/calibration_curve_ensemble_and_all_models_uncalibrated.pdf}
        % \caption{}
        % \label{fig_discussion:calibration_curve_ensemble_and_all_models_uncalibrated}
    \end{subfigure}
    % \hfill
    \begin{subfigure}[c]{0.48\columnwidth}
        \centering
        \includegraphics[width=1\columnwidth]{paper_retrospective_calibration_plots/histogram_ensemble_and_single_model.pdf}
        % \caption{}
        % \label{fig_discussion:histogram_ensemble_and_single_model}
    \end{subfigure}
    \caption[Calibration curve for the uncalibrated stroke recognition model and empirical distribution of predicted probabilities.]{%
        Calibration curve for the uncalibrated stroke recognition ensemble model (left) and the histogram of predicted probabilities (right). 
        We use the ensemble that achieved the median F1-score which is the one also reported in \cref{tab_retrospective:table3-occlusion-analysis} and \cref{fig_retrospective:figure1-roc-curve}.}
    \label{fig_discussion:retrospective-paper-calibration-curve-of-uncalibrated-model}
\end{figure}

It is interesting to note that the ensemble model is better calibrated than the individual models that form it. Since an ensemble output probability is computed as the harmonic mean of the five individual model probabilities, it is easy to see that the ensembles output probability can never exceed the maximum probability predicted among the individual models. This property tends to make ensemble probabilities less extreme and, since the individual models are overconfident, this results in better calibration (see also the histogram in \cref{fig_discussion:retrospective-paper-calibration-curve-of-uncalibrated-model}).

To calibrate the ensemble model, we can use methods such as Platt-scaling \parencite{platt_probabilistic_1999} or isotonic regression \parencite{zadrozny_transforming_2002}. 
In both cases, we fit a simple regression model (logistic or isotonic) to the predicted probabilities and the target labels on some validation set. 
In \cref{fig_discussion:retrospective-paper-calibration-curve-sigmoid-isotonic} we have done so for the ensemble model using our validation set for fitting and visualizing the results on the test set. On the left, we plot the resulting calibration curves and on the right we show the logistic and isotonic fits. We see that both methods result in quite good calibrations\footnote{Test set Brier scores: Uncalibrated = $0.003500$, logistic = $0.001807$, isotonic = $0.001774$. Relative improvement in Brier score compared to uncalibrated (Brier skill score): Logistic = $0.4830$, isotonic = $0.4924$.} and that the predicted probabilities are shifted towards smaller values.

\begin{figure}
    \begin{subfigure}[c]{0.48\columnwidth}
        \centering
        \includegraphics[width=1\columnwidth]{paper_retrospective_calibration_plots/calibration_curves_ensemble.pdf}
        % \caption{}
        % \label{fig_discussion:calibration_curve_ensemble_and_all_models_uncalibrated}
    \end{subfigure}
    % \hfill
    \begin{subfigure}[c]{0.48\columnwidth}
        \centering
        \includegraphics[width=1\columnwidth]{paper_retrospective_calibration_plots/calibration_fits_ensemble.pdf}
        % \caption{}
        % \label{fig_discussion:histogram_ensemble_and_single_model}
    \end{subfigure}
    \caption[Calibration fits and curves for the stroke recognition model using Platt-scaling and isotonic regression for calibration.]{ Calibration curves using sigmoid and isotonic calibration fits for the stroke recognition ensemble model (left) and the calibration fits (right). We use the ensemble that achieved the median F1-score which is the one also reported in \cref{tab_retrospective:table3-occlusion-analysis} and \cref{fig_retrospective:figure1-roc-curve}.}
    \label{fig_discussion:retrospective-paper-calibration-curve-sigmoid-isotonic}
\end{figure}

While correct calibration provides some interpretability to the predictions, the actual values of the probabilities are still tied of the model's performance. For example, with the logistic calibration, we can note that the predicted probabilities have been shifted so far towards smaller values that the highest possible predicted probability is about 50\%. This means strokes can be predicted at best with a 50\% chance of being correct. 
If this model were to be deployed in a randomized controlled trial, the calibrated probabilities might be incorporated in the design of the user interface such that a user can discern between confident and uncertain model predictions. However, it is easy to imagine how this generally low confidence might lead the user to ignore the model. This highlights a problem with uncertainties

@@

>>>>

During the three years since the start of this project, the rapid development of machine learning that started about a decade ago with the 2012 ImageNet competation and the work of \textcite{krizhevsky_imagenet_2012} all but slowed down. 
In this section we will briefly revisit selected chapters of the thesis and provide some additional thoughts and discussion.

\section{Speech representations and uncertainty}
%
In this thesis we have studied two different approaches to learning speech representations: VAEs in \cref{chp:paper-hierarchical,chp:paper-modelagnostic,chp:paper-benchmarking} and self"=supervised methods in \cref{chp:paper-review}. 
We found that the probabilistic formulation of VAEs provide benefits for their application to uncertainty quantification but that they are generally inferior to self"=supervised methods when it comes to performance on downstream tasks, such as phoneme recognition (see e.g. \cref{tab: phoneme recognition (PER), tab:unsupervisedASR}). 
Considering VAEs and self"=supervised methods, it seems we cannot build a model that has a principled probabilistic representation of uncertainty without sacrificing some of the ability to learn rich representations that are useful for downstream applications.

% VAEs are good at uncertainty quantification (let's at least say that they are).
% SSL methods are very good at representation learning as reflected by their dominance on downstream task performance in for text and speech.

@@

One way to improve the representations learned in VAEs is via semi-supervised learning. 
Here a small number of labels are used to inform which patterns that are learned from a large, mostly unlabelled, data set. 
This is usually done by defining a new stochastic variable as the target and deriving a semi-supervised version of the ELBO that accomadates using the labels when they are available, or marginalizing the target variable when they are not. The VAE is then trained on the labelled and unlabelled data simultanously.

VAEs are strong models for semi-supervised learning as proven in several works \parencite{kingma_semi-supervised_2014,kingma_improved_2016,maaloe_biva_2019}. 
However, self"=supervised methods methods have established themselves as superior also in this setting \cite{jiang_speech_2021, liu_learning_2023}. 
Furthermore, by training on labelled and unlabelled data simultaneously, the semi-supervised setting of VAEs is more constraining than it is for self-supervised methods that divide the training into pre"=training and finetuning. Here, expensive pre"=training yields a single foundation model that can be fine"=tuned at relatively low cost for a large number of downstream tasks, while VAEs must learn all relevant downstream tasks while also learning from the unlabelled data, which is more expensive, but also requires retraining on the unlabelled data, when a .

Masked pre"=trained as maximum likelihood

@@

@@

\section{\Cref{chp:paper-retrospective} revisited: Calibration}
Our work on stroke recognition in \cref{chp:paper-retrospective} focuses on the predictive performance of the ensemble model and an analysis of feature importance, but does not explicitly consider uncertainty estimation. 
As we discussed in \cref{subsec:model-calibration}, such a model can be calibrated to predict probabilities that are aligned with the empirical probability of the model being correct on some validation set. Here we present and discuss model calibration for the ensemble model.

We compute the calibration curve by sorting the probabilities predicted on the test set into a number of bins spanning the range from zero to one. For each bin $b$, we compute the mean predicted probability $\bar{p}$ and the fraction of examples for which the model predicted correctly $r_{b}$. The calibration curve is the drawn from the $\{(\bar{p}_{b}, r_{b})\}_b$ pairs. A perfectly calibrated model would have the same fraction of correct predictions in any given bin, as that bin's mean value, $\bar{p}_{b} = r_{b}\,\forall\,b$.

The calibration curve for the uncalibrated stroke recognition ensemble and its constituent models is plotted in \cref{fig_discussion:retrospective-paper-calibration-curve-of-uncalibrated-model} along with a histogram of its predicted probabilities. The miscalibration issue that we previously discussed is clearly visible as a strong overconfidence. 
Additionally, the ensemble model is appreciably better calibrated than its constituent models. Since the ensemble output probability is computed as the harmonic mean of the five constituent model probabilities, the ensemble's output probability can never exceed the maximum probability predicted between the constituent models. This property tends to make ensemble probabilities less extreme and, since the constituent models are overconfident, this results in better calibration (see also the histogram in \cref{fig_discussion:retrospective-paper-calibration-curve-of-uncalibrated-model}).

\begin{figure}
    \begin{subfigure}[c]{0.48\columnwidth}
        \centering
        \includegraphics[width=1\columnwidth]{paper_retrospective_calibration_plots/calibration_curve_ensemble_and_all_models_uncalibrated.pdf}
        % \caption{}
        % \label{fig_discussion:calibration_curve_ensemble_and_all_models_uncalibrated}
    \end{subfigure}
    % \hfill
    \begin{subfigure}[c]{0.48\columnwidth}
        \centering
        \includegraphics[width=1\columnwidth]{paper_retrospective_calibration_plots/histogram_ensemble_and_single_model.pdf}
        % \caption{}
        % \label{fig_discussion:histogram_ensemble_and_single_model}
    \end{subfigure}
    \caption[Calibration curve for the uncalibrated stroke recognition model and empirical distribution of predicted probabilities.]{%
        Calibration curve for the uncalibrated stroke recognition ensemble model (left) and the histogram of predicted probabilities (right) for the test set. 
        We use the ensemble that achieved the median F1-score reported in \cref{tab_retrospective:table3-occlusion-analysis} and \cref{fig_retrospective:figure1-roc-curve}.}
    \label{fig_discussion:retrospective-paper-calibration-curve-of-uncalibrated-model}
\end{figure}

To calibrate the ensemble model, we can use methods such as Platt-scaling \parencite{platt_probabilistic_1999} or isotonic regression \parencite{zadrozny_transforming_2002}. 
We fit a simple regression model (logistic or isotonic) to the predicted probabilities and the target labels on the validation set and use it to adjust the probabilities predicted on the test set. We show the resulting calibration curves on the left in \cref{fig_discussion:retrospective-paper-calibration-curve-sigmoid-isotonic} and the logistic and isotonic fits on the right. 
% In \cref{fig_discussion:retrospective-paper-calibration-curve-sigmoid-isotonic} we have done so for the ensemble model and visualize the results on the test set. 
% On the left, we plot the resulting calibration curves and on the right we show the logistic and isotonic fits. 
We see that both methods result in quite good calibrations\footnote{Brier scores on test set: Uncalibrated = $0.003500$, logistic = $0.001807$, isotonic = $0.001774$. Relative improvement in Brier score compared to uncalibrated (Brier skill score): Logistic = $0.4830$, isotonic = $0.4924$.} and that the predicted probabilities are shifted towards smaller values. Since there are few stroke cases overall, few examples receive a high predicted probability (see histogram in \cref{fig_discussion:retrospective-paper-calibration-curve-of-uncalibrated-model}). This leads to a lack of data for the calibration fits at high predicted probabilities which can be seen to especially affect the nonparametric isotonic regression.

@@

\begin{figure}
    \begin{subfigure}[c]{0.48\columnwidth}
        \centering
        \includegraphics[width=1\columnwidth]{paper_retrospective_calibration_plots/calibration_curves_ensemble.pdf}
        % \caption{}
        % \label{fig_discussion:calibration_curve_ensemble_and_all_models_uncalibrated}
    \end{subfigure}    
    % \hfill
    \begin{subfigure}[c]{0.48\columnwidth}
        \centering
        \includegraphics[width=1\columnwidth]{paper_retrospective_calibration_plots/calibration_fits_ensemble.pdf}
        % \caption{}
        % \label{fig_discussion:histogram_ensemble_and_single_model}
    \end{subfigure}    
    \caption[Calibration fits and curves for the stroke recognition model using Platt-scaling and isotonic regression for calibration.]{ Calibration curves using sigmoid and isotonic calibration fits for the stroke recognition ensemble model (left) and the calibration fits (right) for the test set. We use the ensemble that achieved the median F1-score reported in \cref{tab_retrospective:table3-occlusion-analysis} and \cref{fig_retrospective:figure1-roc-curve}.}
    \label{fig_discussion:retrospective-paper-calibration-curve-sigmoid-isotonic}
\end{figure}

Clinicians in intensive care units and emergency departments have been found to strongly agree that a singular focus on overall accuracy cannot ensure sustained use of and trust in a model \cite{tonekaboni_what_2019}. Clinicians expect an alert to present a prediction that aligns with patient status. Despite expert-agreed thresholds for when alerts should be triggered, however, many alerts may not be aligned since the unbalanced and ambiguous nature of many predictive problems in healthcare can lead to low predictive precision \parencite{umscheid_development_2015, cite14, cite15, wenstrup_retrospective_2023}. This can lead to alarm fatigue \parencite{embi_evaluating_2012} and can undermine the use and endorsement by clinicians of such systems \parencite{guidi_clinician_2015}. 
The expected precision of the stroke recognition model of $24.9\%$ (95\% confidence interval $24.3-25.5\%)$ means it might be wrong three out of four times it predicts a positive. This unfortunately makes alarm fatigue likely among its potential users which might diminish it's real-world impact.

If the stroke recognition model were to be deployed in a randomized controlled trial, such issues might be alleviated by incorporating the calibrated probabilities in the alerts presented to users. With calibrated probabilities, user would be able to discern between certain and uncertain model predictions or might be presented only with predictions that exceed a specific confidence level. Similar approaches have been suggested by clinicians and interviews indicate that predictive uncertainty is perceived as a sort of explanation that complements the prediction \cite{tonekaboni_what_2019}.

% \begin{figure}
%     \centering
%     \includegraphics[width=0.5\textwidth]{paper_retrospective_calibration_plots/agreement_fraction.pdf}
%     \caption[Fractional agreement between stroke recognition model and call-takers as function of predicted probability]{ The fractional agreement between the stroke recognition model and the call-takers as a function of the probability predicted by the model. Both the uncalibrated and the isotonically calibrated model agree more with call-takers towards low and high probabilities. Logistic calibration yielded almost identical results to isotonic.}
%     \label{fig_discussion:retrospective-paper-fractional-agreement-vs-probability}
% \end{figure}

\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]{paper_retrospective_calibration_plots/f1_score_ratio.pdf}
    \caption[]{}
    \label{fig_discussion:retrospective-paper-f1-score-ratio}
\end{figure}

Nevertheless, while correct calibration provides some explanation to the predictions, the actual value of the calibrated probabilities is still tied of the model's performance. By inspecting \cref{fig_discussion:retrospective-paper-calibration-curve-sigmoid-isotonic}, we can see that the logistic calibration shifts the predicted probabilities so far towards zero that the highest possible predicted probability is about 50\%. This means strokes can be predicted at best with a 50\% chance of being correct which limits the potential reduction in alarm fatigue.

\todo[inline]{Finish this sentence and analysis.}
Making matters worse, the most certain model predictions are the ones that agree the most with call-taker judgment (see \cref{fig_discussion:retrospective-paper-fractional-agreement-vs-probability}). Hence, keeping the most certain predictions also means reducing the potential for changing call-taker.. Not really if when the model disagrees at low prob it is always right while when it disagrees at high prob it is always wrong (although unlikely and contrary to calibration - but that was done for the total set of predictions, not subgroups)

This highlights a problem where combining predictions with well-calibrated uncertainties may not be sufficient to ensure sustained use in practice.

### 2023-10-10 chapters/main-contributions.tex
Updates to discussion

\contribitem{chp:paper-hierarchical}
In this work we hypothesize that the likelihood estimate of variational autoencoders is a poor score for out-of-distribution due to on overemphasis on low-level features that generalize between distributions. 
We further hypothesize that a well-formed hierarchy of latent variables provides a tool that can be used to select between which features to emphasize for out-of-distribution detection and hence a way to improve the performance of variational autoencoders on this task. 
We proceed to provide empirical and theoretical evidence that low-level features indeed do dominate the likelihood and propose a new method for out-of-distribution detection using hierarchical variational autoencoders. 
The proposed method computes a likelihood-ratio score for out-of-distribution detection that requires data to be in-distribution across all feature-levels. 
The proposed method is computationally efficient, fully unsupervised, and performs well on several out-of-distribution detection benchmarks. 
% Methods like this are important for the development of safe and reliable machine learning systems, especially in medical applications where the consequences of errors can be significant.

@@

\contribitem{chp:paper-automated}
In this paper, we review the current state-of-the-art in automated medical coding of clinical notes. 
We hypothesize that several previous works underperform for reasons more related to suboptimal hyperparameter tuning, incorrect evaluation, and crude data handling than to model design, and that performance can be improved by addressing these issues. 
With our work, we aim to increase reproducibility and comparability of future work and to provide a baseline for future comparisons. 
We reproduce, analyze and compare several models on the MIMIC-III dataset showing that poor performance is indeed attributable to weak configuration of the training and crudely sampled train-test splits with many extremely rare classes and several without examples in the training data at all. 
We also find a widespread error in the calculation of the macro F1 score. 
To compare models, we propose new data splits created with stratified sampling, use identical experimental setups and tune hyperparameters and decision boundaries. 
By analyzing prediction errors, we confirm the observation of previous work that all models struggle with rare codes, although long documents only have a negligible impact on performance, contrary to previous claims. 
We also present the first comprehensive results on the newly released MIMIC-IV dataset using the reproduced models.

>>>>

\contribitem{chp:paper-hierarchical}
In this work we hypothesize that the likelihood estimate of variational autoencoders is a poor score for out-of-distribution due to an overemphasis on low-level features that generalize between distributions. 
We further hypothesize that a well-formed hierarchy of latent variables provides a tool that can be used to select between which features to emphasize for out-of-distribution detection and hence a way to improve the performance of variational autoencoders on this task. 
We proceed to provide empirical and theoretical evidence that low-level features indeed do dominate the likelihood and propose a new method for out-of-distribution detection using hierarchical variational autoencoders. 
The proposed method computes a likelihood-ratio score for out-of-distribution detection that requires data to be in-distribution across all feature-levels. 
The proposed method is computationally efficient, fully unsupervised, and performs well on several out-of-distribution detection benchmarks. 
% Methods like this are important for the development of safe and reliable machine learning systems, especially in medical applications where the consequences of errors can be significant.

@@

\contribitem{chp:paper-automated}
In this paper, we review the current state-of-the-art in automated medical coding of clinical notes. 
We hypothesize that several previous works underperform for reasons more related to suboptimal hyperparameter tuning, incorrect evaluation, and crude data handling than to model design, and that performance can be improved by addressing these issues. 
With our work, we aim to increase reproducibility and comparability of future work and to provide a baseline for future comparisons. 
We reproduce, analyze and compare several models on the MIMIC-III dataset showing that poor performance is indeed attributable to weak configuration of the training and crudely sampled train-test splits with many extremely rare classes and several without examples in the training data at all. 
We also find a widespread error in the calculation of the macro F1-score. 
To compare models, we propose new data splits created with stratified sampling, use identical experimental setups and tune hyperparameters and decision boundaries. 
By analyzing prediction errors, we confirm the observation of previous work that all models struggle with rare codes, although long documents only have a negligible impact on performance, contrary to previous claims. 
We also present the first comprehensive results on the newly released MIMIC-IV dataset using the reproduced models.
